---
title: "Assessment 0"
output: html_document
---


## Introduction

In the following code we analyse COVID-19 hospitalisation data within the UK and across the world. We aim to look for any relations between the data through the code we find, and seeing how we can improve our findings to give us more insight into how the global pandemic affected healthcare facilities worldwide.


The following pre-requisites are required:
```{r}
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
library(reshape2)
library(lubridate)
```


## Instructions for importing the dataset

1. The first dataset is here: https://ourworldindata.org/covid-hospitalizations
2. You can find a link to the CSV file on Our World in Data's GitHub here, called 'covid-hospitalizations.csv': https://github.com/owid/covid-19-data/tree/master/public/data/hospitalizations
3. Save this file to your desired location, under the file-name 'covid-hospitalizations.csv'



## Exploratory Data Analaysis



### Initial Analaysis

In this section we will complete an first analysis to see how different countries across the world coped.

```{r}
# Read the dataset
data <- read.csv("/Users/harrywilson/Desktop/DataScienceToolbox/DataSets/covid-hospitalizations.csv")

# Select only the relevant columns
data$date <- as.Date(data$date)
icu_data <- data %>%
    filter(indicator == "Daily ICU occupancy") %>%
    select(date, value)  

# Make the plot
ggplot(icu_data, aes(x = date, y = value)) +
    geom_line(color = "blue") +
    labs(title = "Daily ICU Occupancy Over Time",
         x = "Date",
         y = "Daily ICU Occupancy") +
    theme_minimal()
```



##### A brief look

In this first graph, we are looking at all the countries in our data set and their sum of ICU (Intensive Care Unit) occupancies throughout the pandemic and the year that follows. I want to filter our data into only selected countries, as this data only looks at 50 countries, and at the moment we don't even have a list of them. So this isn't too informative aside from looking at general trends across the globe. Which we can see the ICU rates have dropped massively since their last peak in the early months of 2021.

```{r}
library(ggplot2)
library(dplyr)

# Read the dataset
data <- read.csv("/Users/harrywilson/Desktop/DataScienceToolbox/DataSets/covid-hospitalizations.csv")

# Convert date column to Date type
data$date <- as.Date(data$date)

# Select only relevant columns and filter for specific locations
selected_locations <- c("England", "France", "Portugal")  # Replace with your desired countries

icu_data <- data %>%
    filter(indicator == "Daily ICU occupancy", entity %in% selected_locations) %>%
    select(date, entity, value)

# Make the plot
ggplot(icu_data, aes(x = date, y = value, color = entity)) +
    geom_line() +
    labs(title = "Daily ICU Occupancy Over Time",
         x = "Date",
         y = "Daily ICU Occupancy") +
    theme_minimal()

```



#### A more focused look

From this graph, we can see that each ICU occupation rate in each of the countries followed a similar pattern, by looking at the peaks and troughs. However, the problem here is that clearly our different locations have different populations. So it is not really a fair insight into how each country was doing. We can look at this more fairly by instead looking at "ICU Deaths Per Million" of each of these populations.





```{r}
library(ggplot2)
library(dplyr)

# Read the dataset
data <- read.csv("/Users/harrywilson/Desktop/DataScienceToolbox/DataSets/covid-hospitalizations.csv")

# Convert date column to Date type
data$date <- as.Date(data$date)

# Select only relevant columns and filter for specific locations
selected_locations <- c("England", "France", "Portugal")  # Replace with your desired countries

icu_data <- data %>%
    filter(indicator == "Daily ICU occupancy per million", entity %in% selected_locations) %>%
    select(date, entity, value)

# Make the plot
ggplot(icu_data, aes(x = date, y = value, color = entity)) +
    geom_line() +
    labs(title = "Daily ICU Occupancy per Million Over Time",
         x = "Date",
         y = "Daily ICU Occupancy per Million") +
    theme_minimal()

```



#### Beds per million analysis

We can see that each oc the countries, France initially struggled the most, doubliung Englands ICU occupancy per million. This shows how much of a strain they were under, which is backed up by the article [1]. We see England coped okay relatively, however Portugal was also put under a lot of strain in early 2021, again backed up by [2]




### Looking in depth

We will look at the different countries we can analyse in this dataset, along with what there is to analyse in the set.



```{r}
unique_locations <- unique(data$entity)
print(unique_locations)
unique_indicators <- unique(data$indicator)
print(unique_indicators)
```



###### Analysis 2 can be on finding if there is any relation (regression?) between daily icu occupancy and daily hospital occupancy, would expect to find that there is, but not a constant one due to the introduction of vaccines



```{r}
# Assuming your dataset is already loaded as 'data'

# Filter for 'England' and for the relevant indicators
england_data <- data %>%
    filter(entity == "England", 
           indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy"))

# Reshape the data for correlation calculation
england_wide <- england_data %>%
    pivot_wider(names_from = indicator, values_from = value)

# Calculate correlation matrix
cor_matrix <- cor(england_wide[, c("Daily ICU occupancy", "Daily hospital occupancy")], use = "complete.obs")

# Load the necessary libraries
library(ggplot2)
library(reshape2)

# Melt the correlation matrix to long format for heatmap plotting
cor_melt <- melt(cor_matrix)

# Plot heatmap
ggplot(cor_melt, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
    labs(title = "Correlation between Daily ICU and Hospital Occupancy in England",
         fill = "Correlation") +
    theme_minimal()
```


### A first heatmap

From this, we can see there is a clear correlation between Daily ICU Occupancy and Daily Hospital Admissions, which was to be expected. I am now intrigued to see if this relation stays as strong throughout the pandemic within England.


```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)

# Convert date column to Date type if not already done
data$date <- as.Date(data$date)

# Filter the data for 'England' and the relevant indicators
data_filtered <- data %>%
    filter(entity == "England", 
           indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy")) %>%
    select(entity, date, indicator, value)

# Aggregate data by month to reduce overcrowding
data_filtered <- data_filtered %>%
    mutate(month = floor_date(date, "month"))

# Pivot data to wide format to have ICU and hospital occupancy in separate columns
data_wide <- data_filtered %>%
    pivot_wider(names_from = indicator, values_from = value)

# Remove rows with missing values
data_wide <- na.omit(data_wide)

# Calculate the correlation between ICU occupancy and hospital occupancy for each month
monthly_corr <- data_wide %>%
    group_by(month) %>%
    summarise(correlation = cor(`Daily ICU occupancy`, `Daily hospital occupancy`, use = "complete.obs"))

# Create the heatmap
ggplot(monthly_corr, aes(x = month, y = 1, fill = correlation)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Correlation") +
    labs(title = "Monthly Correlation Between ICU and Hospital Occupancy in England",
         x = "Month",
         y = "",
         fill = "Correlation") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_text(size = 8, angle = 45, hjust = 1),  # Smaller x-axis text
          plot.title = element_text(size = 12)) +  # Adjust title size
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    scale_y_discrete(limits = c("Correlation"))  # Ensure y-axis is labeled properly

```
#### Inspecting correlation

From this, we see that there is a clear correlation throughout the pandemic, however it looks like there is no correlation in one of the months. We assume this may be because of missing data, so we adjust our code to allow for all missing data to see if this is the problem.


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)

# Convert date column to Date type if not already done
data$date <- as.Date(data$date)

# Filter the data for 'England' and the relevant indicators
data_filtered <- data %>%
    filter(entity == "England", 
           indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy")) %>%
    select(entity, date, indicator, value)

# Aggregate data by month
data_filtered <- data_filtered %>%
    mutate(month = floor_date(date, "month"))

# Pivot data to wide format
data_wide <- data_filtered %>%
    pivot_wider(names_from = indicator, values_from = value)

# Remove rows with missing values
data_wide <- na.omit(data_wide)

# Calculate the correlation
monthly_corr <- data_wide %>%
    group_by(month) %>%
    summarise(correlation = cor(`Daily ICU occupancy`, `Daily hospital occupancy`, use = "complete.obs"), .groups = "drop")

# Create a complete sequence of months
all_months <- data.frame(month = seq(min(data_filtered$month), max(data_filtered$month), by = "month"))

# Join with monthly correlation to fill in missing months
monthly_corr <- all_months %>%
    left_join(monthly_corr, by = "month") %>%
    replace_na(list(correlation = NA))  # or set to 0 if you prefer

# Create the heatmap
ggplot(monthly_corr, aes(x = month, y = 1, fill = correlation)) +
    geom_tile(color = "white", width = 25) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Correlation") +
    labs(title = "Monthly Correlation Between ICU and Hospital Occupancy in England",
         x = "Month",
         y = "",
         fill = "Correlation") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
          plot.title = element_text(size = 12)) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month")

```

We see from this heat map that a lack of data points wasn't the problem, and values were taken in the middle. An explanation for this is possibly the vaccine starting to take effect so the ICU admissions decreasing, however similar amounts of people are still being admitted to hospital, until it slowly averages out.


###### Analaysis 3 could be looking at ICU rates in different countries, is there any relation? Are our findings supported by any other claims. I would expect the west to have had better resources so have done better


##### Need a simple linear regression model


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter the dataset for 'England' and the relevant indicators (ICU and hospital occupancy)
england_data <- data %>%
  filter(entity == "England" & indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy"))

# Pivot data to wide format to have ICU and hospital occupancy in separate columns
england_wide <- england_data %>%
  pivot_wider(names_from = indicator, values_from = value)

# Remove rows with missing values (NA)
england_wide <- na.omit(england_wide)

# Perform linear regression: Hospital Occupancy as a function of ICU Occupancy
lm_model <- lm(`Daily hospital occupancy` ~ `Daily ICU occupancy`, data = england_wide)

# Print the summary of the linear regression model
summary(lm_model)

# Scatter plot with regression line
ggplot(england_wide, aes(x = `Daily ICU occupancy`, y = `Daily hospital occupancy`)) +
  geom_point(aes(color = "ICU vs Hospital Occupancy"), size = 2) +  # Scatter plot
  geom_smooth(method = "lm", se = FALSE, color = "blue", size = 1.5) +  # Regression line
  labs(title = "Linear Regression: Daily Hospital Occupancy vs ICU Occupancy in England",
       x = "Daily ICU Occupancy",
       y = "Daily Hospital Occupancy") +
  theme_minimal() +
  scale_color_manual(name = "Occupancy Type", values = c("ICU vs Hospital Occupancy" = "green")) +
  theme(legend.position = "none")

```
Here we can see a pretty good representation for Daily ICU occupancy to Daily Hospital Occupancy with this linear regression model, however we are going to try and change the colour gradient over time of the points to see if there are any points which look like they could be better represented by another regression model. 
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming your dataset 'data' is already loaded
# and contains columns: 'entity', 'date', 'indicator', and 'value'

# Convert date column to Date type if not already done
data$date <- as.Date(data$date)

# Filter the data for 'England' and the relevant indicators
data_filtered <- data %>%
    filter(entity == "England", 
           indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy")) %>%
    select(entity, date, indicator, value)

# Pivot data to wide format to have ICU and hospital occupancy in separate columns
data_wide <- data_filtered %>%
    pivot_wider(names_from = indicator, values_from = value)

# Remove rows with missing values
data_wide <- na.omit(data_wide)

# Plot the scatter plot with color representing date
ggplot(data_wide, aes(x = `Daily ICU occupancy`, y = `Daily hospital occupancy`, color = date)) +
    geom_point() +
    scale_color_gradient(low = "blue", high = "red", 
                         name = "Date Progression",
                         breaks = as.Date(c("2020-01-06", "2021-01-01", "2022-01-01", "2023-01-01", "2024-08-12")),
                         labels = c("Jan 2020", "Jan 2021", "Jan 2022", "Jan 2023", "Aug 2024")) +
    labs(title = "ICU vs Hospital Occupancy in England Over Time",
         x = "Daily ICU Occupancy",
         y = "Daily Hospital Occupancy") +
    theme_minimal()

```
from this colour gradient, it looks like towards the back end of our data, and the start of our data, we have a better correlation, so it would better to see each of them individually. This is because one is before the vaccine, and the others is after the vaccine has taken full effect.


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Convert date column to Date type if not already done
data$date <- as.Date(data$date)

# Filter the data for 'England', relevant indicators, and the first 6 months
data_filtered <- data %>%
    filter(entity == "England", 
           indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy"),
           date >= as.Date("2020-01-01") & date <= as.Date("2020-06-30")) %>%
    select(entity, date, indicator, value)

# Pivot data to wide format to have ICU and hospital occupancy in separate columns
data_wide <- data_filtered %>%
    pivot_wider(names_from = indicator, values_from = value)

# Remove rows with missing values
data_wide <- na.omit(data_wide)

# Perform linear regression
model <- lm(`Daily hospital occupancy` ~ `Daily ICU occupancy`, data = data_wide)

# Plot the scatter plot with regression line (green dots and blue straight line)
ggplot(data_wide, aes(x = `Daily ICU occupancy`, y = `Daily hospital occupancy`)) +
    geom_point(color = "green") +  # Green dots
    geom_abline(intercept = coef(model)[1], slope = coef(model)[2], color = "blue", linetype = "solid") +  # Blue straight regression line
    labs(title = "Linear Regression of ICU vs Hospital Occupancy (Jan 2020 - June 2020)",
         x = "Daily ICU Occupancy",
         y = "Daily Hospital Occupancy") +
    theme_minimal()

```

##### Here, we are looking at the rates from January 2020 up to June 2020, so at the start of the pandemic, with no vaccine available. Great line of best fit.

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Convert date column to Date type if not already done
data$date <- as.Date(data$date)

# Filter the data for 'England', relevant indicators, and the first 6 months
data_filtered <- data %>%
    filter(entity == "England", 
           indicator %in% c("Daily ICU occupancy", "Daily hospital occupancy"),
           date >= as.Date("2023-01-01") & date <= as.Date("2023-06-30")) %>%
    select(entity, date, indicator, value)

# Pivot data to wide format to have ICU and hospital occupancy in separate columns
data_wide <- data_filtered %>%
    pivot_wider(names_from = indicator, values_from = value)

# Remove rows with missing values
data_wide <- na.omit(data_wide)

# Perform linear regression
model <- lm(`Daily hospital occupancy` ~ `Daily ICU occupancy`, data = data_wide)

# Plot the scatter plot with regression line (green dots and blue straight line)
ggplot(data_wide, aes(x = `Daily ICU occupancy`, y = `Daily hospital occupancy`)) +
    geom_point(color = "green") +  # Green dots
    geom_abline(intercept = coef(model)[1], slope = coef(model)[2], color = "blue", linetype = "solid") +  # Blue straight regression line
    labs(title = "Linear Regression of ICU vs Hospital Occupancy (Jan 2023 - Jun 2023)",
         x = "Daily ICU Occupancy",
         y = "Daily Hospital Occupancy") +
    theme_minimal()

```

##### From looking at this, it seems like we have a good regression model for ICU Occupancy to Hospital Occupancy from January 2023 to June 2023, after the heights of the COVID pandemic