## Data Import 

To automate this, DIFFERENTLY AND BETTER THAN THE RECORDING, we will use internal R functions which work regardless of file system. We would like a cross-platform (Windows/Linux/Mac) solution, and this is provided by the `file.path` function in R.

We want to obtain the data here:

https://github.com/owid/covid-19-data/tree/master/public/data/hospitalizations



```{r}
# Use the full path to the CSV file directly
csv_file_path <- "/Users/harrywilson/Desktop/DataScienceToolbox/DataSets/covid-hospitalizations.csv"

# Now read the CSV file
if (file.exists(csv_file_path)) {
    data <- read.csv(csv_file_path)
    head(data)
} else {
    print("File not found")
}

```

```{r}
str(data)
```
```{r}
# Print unique location names in the dataset
unique_locations <- unique(data$entity)
print(unique_locations)
```





```


```{python}
import pandas as pd


# Update the file path
df = pd.read_csv(r"/Users/harrywilson/Desktop/DataScienceToolbox/DataSets/covid-hospitalizations.csv")

# Print the DataFrame
print(df)


```


