---
title: "R_research"
author: "Mark Milner"
date: "2024-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Packages 

```{r}
install.packages("ggplot2")  # For visualization

```

## Obtaining the data 

Datasets at the following links:

https://github.com/owid/covid-19-data/tree/master/public/data

https://covid19datahub.io/articles/data.html#download-all-in-one-1 
- https://covid19datahub.io/articles/r.html?q=r#quickstart
- https://github.com/covid19datahub/COVID19?tab=readme-ov-file

https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports_us
- https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series


```{r}

data_path = "C:/Users/markm/OneDrive/Documents/University/Year 4/dst/data"

our_world_in_data_path = file.path(data_path, "our_world_in_data_covid_data.csv")

owid = read.csv(our_world_in_data_path)

```

```{r}
head(owid)
```

```{r}
# Filter for the United States
# Load the dplyr package for data manipulation
library(dplyr)
colnames(owid)
uk_data = filter(owid, location =="United Kingdom")
head(uk_data)

```


```{r}

```


```{r}
#unlink("C:/Users/markm/AppData/Local/R/win-library/4.3/00LOCK", recursive = TRUE)
install.packages("COVID19")
```

```{r}
library("COVID19")
x <- covid19()
```
```{r}
#head(x)
#1: country-level data  
#2: state-level data
#3: lower-level data
#x = covid19(country = 'US',level=2)
#head(x)
metadata = covid19('metadata')
head(metdata)
```

## Exploratory Data Analysis.

### Heading 1

### Heading 2 

### Heading 3

### Heading 4

### Conclusions / answering the questions 

• What are the broad types of data?
• What are the main types of resource?
• What type of problems can the resources solve?
• Are there any generic data science resources that might be applicable? In
what sense are they applicable?
• How might the approach be compared to other approaches, and/or applied
across different datasets?
• How is the experience of sharing code via GitHub limiting, and/or enabling?










### Bibl

https://www.rdocumentation.org/packages/dplyr/versions/1.0.10/topics/filter


